Nonlinear Intrinsic Variables and State Reconstruction in Multiscale Simulations

### 2013-12-27 paper/draft6.tex
Added files related to paper

@@

\title{Intrinsic Variables and State Reconstruction in Multiscale Simulations}

@@

\author{Neta Rabin}
\email{netar@afeka.ac.ilil}
\affiliation{Department of Exact Sciences, Afeka Tel-Aviv Academic College of Engineering, Tel-Aviv, Israel}

@@

\begin{abstract}
Finding informative low-dimensional descriptions of high-dimensional simulation data
(like the ones arising in molecular dynamics or kinetic Monte Carlo simulations of
physical and chemical processes) is crucial to understanding physical phenomena, and can
also dramatically assist in accelerating the simulations themselves.
%
In this paper, we discuss and illustrate the use of nonlinear intrinsic variables (NIV)
in the mining of high-dimensional multiscale simulation data.
%
In particular, we focus on the way NIV allows us to functionally merge different
simulation ensembles, and {\em different partial observations of these ensembles}, as well
as to infer variables not explicitly measured.
%
The approach relies on certain simple features of the underlying process variability to
filter out measurement noise and systematically recover a unique reference coordinate frame.
%
We illustrate the approach through two distinct sets of multiscale {\color{red} atomistic} simulations:
a stochastic simulation of an enzyme reaction network exhibiting both fast and slow time scales,
 and a molecular dynamics simulation of alanine dipeptide in explicit water.

@@

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data {\color{red} TODO: add references}.
%
These developments have the potential to significantly enable the computational exploration
of physicochemical problems.
%
If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
molecular dynamics simulation of a macromolecule in solution or the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.

While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
 data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis \cite{jolliffe2005principal},
different data sets on the same low-dimensional hyperplane in the ambient space
will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
%
While this can be easily rectified by a linear transformation, the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data sets are obtained using different instrumental modalities
(such as when one wants to merge molecular dynamics data with, for example, spectral information {\color{red} TODO: add specific example of spectral information}).
%
Clearly, the ability to systematically construct a {\em unique, consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities -
what we call Nonlinear Intrinsic Variables - is invaluable.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements through the construction
of nonlinear observers of unmeasured variables.
%
This empirical mapping also requires good interpolation tools in embedding space; to this
end, here we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.

We will demonstrate our methodologies with two distinct illustrative examples.
%
The first is a simulation
of two Goldbeter-Koshland modules in an enzyme kinetics model using Gillespie's Stochastic Simulation
Algorithm (SSA) \cite{gillespie1977exact};
in certain parameter regimes, separation of time scales is known
to reduce this model to an effective two-dimensional description \cite{zagaris2012stability}.
%
Although this example is rather simple, it will serve as an introduction to our techniques and highlight the main features of the algorithms.
%
The second example is a molecular dynamics
simulation (in explicit water) of a simple peptide fragment (alanine dipeptide) whose folding
dynamics are known to be described through a small set of physical observables \cite{bolhuis2000reaction}.
%
This example will demonstrate how our techniques can have advantages over more common techniques,
such as diffusion maps for dimensionality reduction
and nearest neighbor interpolation for reconstruction.
%
The paper is structured as follows: in Section \ref{sec:NIV} we present the Nonlinear Intrinsic Variable formulation and
the associated inference method.
%
Section \ref{sec:LapPyr} contains our discussion of Laplacian Pyramids that
will be used for the completion of partial observations.
%
Section \ref{sec:examples} contain the results
of applying the approach to simulation data from our two illustrative examples.
%
We conclude with
a summary and discussion of open issues in Section \ref{sec:conclusions}.

@@

\subsection{Overview}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
%
The measured process is assumed (and this is a critical assumption) to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t))
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear, $d \ll n$) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
%
The dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises (the independence is another critical assumption).

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$.
%
By relying on the fact that the empirical method we will describe is independent of the observation function $f$,
we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
%
The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $f$ in various observable domains,
or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $f$ (see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).

The presented method consists of the following main principles.
%
(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise, and hence, creates a ``sphere" of samples in the underlying domain.
%
This sphere is mapped to an ellipse in the observable domain by the measurement function $f$.
%
In this work, the identification of the associated ellipse of samples according to the time labels enables us
to estimate the tangent planes of the observable manifolds {\em via the principal components of the covariance matrices of the samples in these ellipses}.
%
(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be {\em locally invariant to the measurement function $f$}.
%
(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using this Riemannian metric.

\begin{figure*}[ht]
\includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $f$. (Bottom) The underlying variables in which the noises are independent with unit variance. The circle illustrates samples from, say, a short trajectory in time that sample a disc on the manifold. (Top Left) The first set of observed variables. The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation function. In this sketch, we illustrate that the observations might be {\em partial}, i.e., might consist of merely a subset of the observed domain variables. (Top Right) Second set of observable variables. The ellipse illustrates the mapping of the sphere of the underlying samples into this (different) observable domain via a second observation function.}
\label{fig:IntrinsicIllustration}
\end{figure*}

@@

Let $\mathbf{C}(t)$ be the covariance matrix associated with the measured sample $\mathbf{Y}(t)$. In practice, the covariance matrix can be estimated from a short trajectory of samples in time around the sample $\mathbf{Y}(t)$ by
\begin{equation}
	\widehat{\mathbf{C}}(t) = \sum \limits _{\tau = t-L}^{t+L} (\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))(\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))^T
	\label{eq:cov}
\end{equation}
where $\widehat{\boldsymbol{\mu}}(t)$ is the empirical mean of the samples.
%
We define a Riemannian metric between a pair of samples using the associated covariance matrices as
\begin{equation}
	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1}(\mathbf{Y}(t) - \mathbf{Y}(\tau));
	\label{eq:mahalanobis}
\end{equation}
this is the Mahalanobis distance.
%
As previously described, the covariance matrices convey the local variability of the measurements and are utilized to explore and learn the tangent planes of the observable manifold.
%
This information is then utilized in \eqref{eq:mahalanobis} to compare a pair of points according to the directions of their respective tangent planes.
%
The Mahalanobis distance is invariant under linear transformations.
%
Thus, by assuming that the observation function $f$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $f(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to second order, i.e.,
\begin{equation}
	\| \mathbf{x}(t) - \mathbf{x}(\tau) \|^2 = d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) + \mathcal{O}(\| \mathbf{Y}(t) - \mathbf{Y}(\tau)\|^4).
\end{equation}
%
This result implies that the Mahalanobis distance is invariant to the measurement function $f$, and hence,
it yields the same distances between samples obtained under different observation functions or even partial observations.
%
We would like to note that, in general, $f$ being bi-Lipschitz implies that $f$ is invertible.
%
However, in practice, determining whether $f$ contains sufficient information and is ``rich enough'' to completely determine the underlying process is a non-trivial task.

@@

Let $\mathbf{W}$ be a pairwise affinity matrix (kernel) based on a Gaussian, whose $(t,\tau)$-th element is given by
\begin{equation}
	W_{t,\tau} = \exp \left\{ - \frac{ d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) )} {\varepsilon}\right\}
	\label{eq:kernel}
\end{equation}
where $\varepsilon$ is the kernel scale, which can be set according to References \onlinecite{coifman2008graph, hein2005intrinsic}.
%
Based on the kernel, we form a weighted graph, where the measurements $\mathbf{Y}(t)$ are the graph nodes and the weight of the edge connecting node $\mathbf{Y}(t)$ to node $\mathbf{Y}(\tau)$ is $W_{t,\tau}$.
%
In particular, a Gaussian kernel exhibits a notion of locality by defining a neighborhood around each measurement $\mathbf{Y}(t)$ of radius $\varepsilon$,
i.e., measurements $\mathbf{Y}(\tau)$ such that $d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) ) > \varepsilon$ are weakly connected to $\mathbf{Y}(t)$.
%
In practice, we set $\varepsilon$ to be the median of the elements of the kernel matrix.
%
According to the graph interpretation, this implies a well connected graph because each measurement is effectively connected to half of the other measurements \cite{rohrdanz2011determination}.

Let $\mathbf{D}$ be a diagonal matrix whose elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
be a normalized kernel that shares its eigenvectors with the normalized graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}} \:$ \cite{chung1997spectral}.
%
The eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, denoted $\psi_j$, reveal the underlying structure of the data \cite{coifman2005geometric}.
%
Specifically, the $i$-th coordinate of the $j$-th eigenvector can be associated with an intrinsic coordinate $j$ of the sample $\mathbf{x}(i)$ of the underlying process.
%
The eigenvectors are ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$, where $\lambda_j$ is the eigenvalue associated with eigenvector $\psi_j$.
%
Because $\mathbf{W}^{\mathrm{norm}} \sim \mathbf{D}^{-1}\mathbf{W} $, and $\mathbf{D}^{-1}\mathbf{W}$ is row-stochastic,
$\lambda_1 = 1$ and $\psi_1$ is the diagonal of $\mathbf{D}^{1/2}$.
%
The next few eigenvectors describe the geometry of the underlying manifold.
%
However, some eigenvectors can be higher harmonics of the same principal direction (similar to how the eigenfunctions $\cos x$ and $\cos 2x$ both characterize the $x$ direction),
and so one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
%
The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.

Each eigenvector then describes an intrinsic variable for the data set of interest.
%
Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets so that the resulting embeddings are consistent.
%
We first scale the eigenvectors so that the norm of the each eigenvector is equal to the cardinality of the data set.
%
Intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
%
Still, the embedding eigenvectors for two identical data sets could differ by a sign.
%
Reconciling the signs for the embeddings of different data sets can be done in several ways and is somewhat problem-specific.
%
For example, if the mean of the embedding is sufficiently far from 0, the signs can be chosen such that the mean of each eigenvector is greater than 0;
alternatively, we can construct the data sets such that there is a common region between both data sets, and the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of the common regions.

@@

\item
Apply EVD on the normalized kernel and view the values of its principal eigenvectors as the intrinsic variables of the given observations.

@@

\section{Laplacian Pyramids} \label{sec:LapPyr}

In this work, we are not only interested in extracting the underlying variables $\mathbf{x}$ from some (partial) observations $\mathbf{Y}$,
but also interested in extending high-dimensional functions on a set of points which lie in a low-dimensional space - predicting $\mathbf{Y}$ for new points $\mathbf{x}$.
%
Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
%
The algorithm uses Laplacian kernels of decreasing widths to create multiscale representations of $f$;
these representations can be easily extended to new data points.
%
This type of multiscale representation was introduced by Burt and Adelson \cite{burt1983laplacian} for image coding,
and was later shown to be a tight frame by Do and Veterli \cite{do2003framing}.
%
Recently, LP was used to extend nonlinear embedding coordinates to new high-dimensional data points \cite{rabin2012heterogeneous}.
%
We will first review the LP algorithm for approximating and extending a one-dimensional function,
and then describe the application of LP in extending high-dimensional functions.

Let $f: \Gamma \rightarrow \mathbb{R}^n$ be a function that is known on a subset of points $S \subset \Gamma$.
%
A coarse representation of $f$ is generated using a coarse smoothing operator $P_0$.
%
The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(i, j)= s_0^{-1}(t)w_0(i, j)$,
where $w_0(i, j)=e^{-d^2(i, j) / \sigma_0}$ and $s_0(t)=\sum_{j \in S}w_0(i, j)$ is the normalizing term.
%
The pairwise distance $d(i, j)$ is typically the Euclidean distance, and the parameter $\sigma_0$ is set to be large compared to the values of $d^2(i, j)$.
%
The application of $P_0$ to $f$ yields a coarse representation of the function, which we denote by $f_0=P_0(f)$.

>>>>

\makeatletter
\renewcommand\@make@capt@title[2]{%
 \@ifx@empty\float@link{\@firstofone}{\expandafter\href\expandafter{\float@link}}%
  {\textbf{#1}}\@caption@fignum@sep#2\quad
}%
\makeatother

@@

\title{Nonlinear Intrinsic Variables and State Reconstruction in Multiscale Simulations}

@@

\author{Neta Rabin}
\email{netar@afeka.ac.il}
\affiliation{Department of Exact Sciences, Afeka Tel-Aviv Academic College of Engineering, Tel-Aviv, Israel}

@@

\begin{abstract}
Finding informative low-dimensional descriptions of high-dimensional simulation data
(like the ones arising in molecular dynamics or kinetic Monte Carlo simulations of
physical and chemical processes) is crucial to understanding physical phenomena, and can
also dramatically assist in accelerating the simulations themselves.
%
In this paper, we discuss and illustrate the use of nonlinear intrinsic variables (NIV)
in the mining of high-dimensional multiscale simulation data.
%
In particular, we focus on the way NIV allows us to functionally merge different
simulation ensembles, and {\em different partial observations of these ensembles}, as well
as to infer variables not explicitly measured.
%
The approach relies on certain simple features of the underlying process variability to
filter out measurement noise and systematically recover a unique reference coordinate frame.
%
We illustrate the approach through two distinct sets of atomistic simulations:
a stochastic simulation of an enzyme reaction network exhibiting both fast and slow time scales,
 and a molecular dynamics simulation of alanine dipeptide in explicit water.

@@

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data \cite{Tenenbaum2000,Roweis2000,Donoho2003,Belkin2003,Coifman2006}.
%
These developments have the potential to significantly enable the computational exploration
of physicochemical problems.
%
If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.

While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
 data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis \cite{jolliffe2005principal},
different data sets on the same low-dimensional hyperplane in the ambient space
will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
%
While this can be easily rectified by an affine transformation
(see, by analogy, the discussion in \cite{lafon2006data}),
the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data sets are obtained using different instrumental modalities
(such as when one wants to merge molecular dynamics data with, for example, spectral information {\color{red} TODO: add specific example of spectral information}).
%
Clearly, the ability to systematically construct a {\em unique, consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities -
what we will call Nonlinear Intrinsic Variables - is invaluable.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements through the construction
of nonlinear observers of unmeasured variables.
%
%%%RONEN
% we need in the sentence above to mention that we have the analog of a ``training set"
%
To construct this empirical mapping and the associated observers,
accurate interpolation tools must be available in the embedding space; to this
end, here we chose to demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
%%
%%RONEN
%% remember to add all the other options, like kriging..... in the discussion
%%

We will demonstrate our methodologies with two distinct illustrative examples.
%
The first is a simulation
of two Goldbeter-Koshland modules in an enzyme kinetics model using Gillespie's Stochastic Simulation
Algorithm (SSA) \cite{gillespie1977exact};
in certain parameter regimes, separation of time scales is known
to reduce the ODE model of this kinetic scheme to an effective two-dimensional description \cite{zagaris2012stability}.
%
Although this example is rather simple, it will serve as an introduction to our techniques and highlight the main features of the algorithms.
%
The second example is a molecular dynamics
simulation (in explicit water) of a simple peptide fragment (alanine dipeptide) whose folding
dynamics are known to be described through a small set of physical observables \cite{bolhuis2000reaction}.
%
This example will allow us to compate our approach to more common techniques,
such as diffusion maps for dimensionality reduction
and nearest neighbor interpolation for observation reconstrcution.
%
The remainder of the paper is structured as follows: in Section \ref{sec:NIV} we present the Nonlinear Intrinsic Variable formulation and
the associated inference method.
%
Section \ref{sec:LapPyr} contains our discussion of Laplacian Pyramids that
will be used for the completion of partial observations.
%
In section \ref{sec:examples} the results
of the application of the approach to simulation data from our two illustrative examples is presented and discussed.
%
We conclude with a summary and our perspective on open issues in Section \ref{sec:conclusions}.

@@

\subsection{Overview}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
%
The measured process is assumed (and this is a critical assumption) to be a manifestation, in an observable domain, of a low-dimensional diffusion process. Thus, it can be expressed as
\begin{equation}
	\mathbf{Y}(t) = \mathbf{f}(\mathbf{x}(t))
\end{equation}
where $\mathbf{f}:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear, $d \ll n$) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
%
The dynamics of the diffusion process in terms of its underlying variables is described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises (this independence is our second critical assumption).

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$.
%
By relying on the fact that the empirical method we will describe is independent of the observation function $\mathbf{f}$,
we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
%
The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $\mathbf{f}$ in various observable domains,
or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $\mathbf{f}$ 
(see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
%
We remark that the available data should be rich enough, i.e., consists of a sufficient amount of ``historical" data with sufficient variability, in order to get a full empirical model.

The method consists of the following main principles.
%
(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise,
and hence, creates a ``sphere" of samples in the underlying domain.
%%
%%%RONEN
% whether here or in the discussion there should be something about testing this
% practically
%
This sphere is mapped to an ellipse in the observable domain by the measurement function $\mathbf{f}$.
%
In this work, the identification of the associated ellipse of samples according to the time labels enables us
to estimate the tangent planes of the observable manifolds {\em via the principal components of the covariance matrices of the samples in these ellipses}.
%
(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be {\em locally invariant to the measurement function $\mathbf{f}$}.
%
(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using this Riemannian metric.

\begin{figure*}[ht]
\includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $\mathbf{f}$. (Bottom) The underlying variables in which the noises are independent with unit variance. The circle illustrates samples from, say, a short trajectory in time that sample a disc on the manifold. (Top Left) The first set of observed variables. The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation function. In this sketch, we illustrate that the observations might be {\em partial}, i.e., might consist of merely a subset of the observed domain variables. (Top Right) Second set of observable variables. The ellipse illustrates the mapping of the sphere of the underlying samples into this (different) observable domain via a second observation function.}
\label{fig:IntrinsicIllustration}
\end{figure*}

@@

Let $\mathbf{C}(t)$ be the covariance matrix associated with the measured sample $\mathbf{Y}(t)$. In practice, the covariance matrix can be estimated from a short trajectory of samples in time around the sample $\mathbf{Y}(t)$ by
\begin{equation}
	\widehat{\mathbf{C}}(t) = \sum \limits _{\tau = t-L}^{t+L} (\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))(\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))^T
	\label{eq:cov}
\end{equation}
where $\widehat{\boldsymbol{\mu}}(t)$ is the empirical mean of the samples.
%
We define a Riemannian metric between a pair of samples using the associated covariance matrices as
\begin{equation}
	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1}(\mathbf{Y}(t) - \mathbf{Y}(\tau));
	\label{eq:mahalanobis}
\end{equation}
this is the Mahalanobis distance.
%
As previously described, the covariance matrices convey the local variability of the measurements and are utilized to explore and learn the tangent planes of the observable manifold.
%
This information is then utilized in \eqref{eq:mahalanobis} to compare a pair of points according to the directions of their respective tangent planes.
%
The Mahalanobis distance is invariant under affine transformations.
%
Thus, by assuming that the observation function $\mathbf{f}$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $\mathbf{f}(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to second order, i.e.,
\begin{equation}
	\| \mathbf{x}(t) - \mathbf{x}(\tau) \|^2 = d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) + \mathcal{O}(\| \mathbf{Y}(t) - \mathbf{Y}(\tau)\|^4).
\end{equation}
%
This result implies that the Mahalanobis distance is invariant to the measurement function $\mathbf{f}$, and hence,
it yields the same distances between samples obtained under different observation functions or even partial observations.
%
We would like to note that, in general, $\mathbf{f}$ being bi-Lipschitz implies that $\mathbf{f}$ is invertible.
%%%
%%%RONEN
% this is nontrivial and it requires discussion
%  about what manifold thing need to live on for this to be doable
%
However, in practice, determining whether $\mathbf{f}$ contains sufficient information and is ``rich enough'' to completely determine the underlying process is a non-trivial task.
%
In this work, we exploit the fact that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$, which implies that $\mathbf{C}(t)$ is an $n \times n$ positive semidefinite matrix of rank $d$, to empirically infer the dimension $d$. 
%
According to the spectrum of the local covariance matrices and their corresponding spectral gaps, we approximate the rank of the matrices. 
%
Consistent rank estimates among the covariance matrices may imply that the measurements are ``rich enough", and hence, may be good indicators for the dimension $d$. 
%
In addition, since usually the dimension $d$ of the underlying process is smaller than the dimension of the measured process $n$, the covariance matrix is singular and non-invertible. 
%
Thus, in practice we use the pseudo-inverse to compute the inverse matrices in \eqref{eq:mahalanobis}.

@@

Let $\mathbf{W}$ be a pairwise affinity matrix (kernel) based on a Gaussian, whose $(t,\tau)$-th element is given by
\begin{equation}
	W_{t,\tau} = \exp \left\{ - \frac{ d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) )} {\varepsilon}\right\}
	\label{eq:kernel}
\end{equation}
where $\varepsilon$ is the kernel scale, which can be set according to References \onlinecite{coifman2008graph, hein2005intrinsic}.
%
Based on the kernel, we form a weighted graph, where the measurements $\mathbf{Y}(t)$ are the graph nodes and the weight of the edge connecting node $\mathbf{Y}(t)$ to node $\mathbf{Y}(\tau)$ is $W_{t,\tau}$.
%
In particular, a Gaussian kernel exhibits a notion of locality by defining a neighborhood around each measurement $\mathbf{Y}(t)$ of radius $\varepsilon$,
i.e., measurements $\mathbf{Y}(\tau)$ such that $d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) ) > \varepsilon$ are weakly connected to $\mathbf{Y}(t)$.
%
In practice, we set $\varepsilon$ to be the median of the elements of the kernel matrix.
%%%RONEN
%   this is also not really clear, and "effectively connected to half" is
%   also not true -- just need to be said carefully
%
According to the graph interpretation, this implies a well connected graph because each measurement is effectively connected to half of the other measurements \cite{rohrdanz2011determination}.

Let $\mathbf{D}$ be a diagonal matrix whose elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
be a normalized kernel that shares its eigenvectors with the normalized graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}} \:$ \cite{chung1997spectral}.
%
The eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, denoted $\psi_j$, reveal the underlying structure of the data \cite{coifman2005geometric}.
%
Specifically, the $i$-th coordinate of the $j$-th eigenvector can be associated with an intrinsic coordinate $j$ of the sample $\mathbf{x}(i)$ of the underlying process.
%
The eigenvectors are ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$, where $\lambda_j$ is the eigenvalue associated with eigenvector $\psi_j$.
%
Because $\mathbf{W}^{\mathrm{norm}} \sim \mathbf{D}^{-1}\mathbf{W} $, and $\mathbf{D}^{-1}\mathbf{W}$ is row-stochastic,
$\lambda_1 = 1$ and $\psi_1$ is the diagonal of $\mathbf{D}^{1/2}$.
%
The next few eigenvectors can be argued to describe the geometry of the underlying manifold \cite{coifman2005geometric}
%
However, some eigenvectors can be higher harmonics of the same principal direction along the data manifold.
This is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
and with no flux boundary conditions are one-to-one with the values of $x$ for $0 \le x \le 1$);
so one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
%
The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.

Ignoring these ``higher harmonics", each retained eigenvector then describes an intrinsic variable for the data set of interest.
%
Because eigenvectors can be determined up to a scaling factor,
we must scale the eigenvectors from different data sets so that the resulting embeddings are consistent.
%
We first scale the eigenvectors so that $\|\psi_i\| = T$, where $T$ is the number of data points;
intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
%
Still, the computed embedding eigenvectors, even for two identical data sets, may differ by a sign.
%
Reconciling the signs for the embeddings of different data sets can be rationally done in several ways and is somewhat problem-specific.
%
For example, if the mean of the embedding is sufficiently far from 0, we can require $\langle \psi_i \rangle > 0$;
%the signs can be chosen such that the mean of each eigenvector has a prescribed sign;
alternatively, if there is a common region sampled by both data sets, the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of the common region data.
%
We will return to the issue of embedding consistency for different data sets in our concluding discussion; for the moment,
we will assume that our different sets sample the same region of data space in a representative enough way for the
correspondence between the sequences of retained eigenvectors for different embeddings to be obvious.

@@

\item
Apply eigenvector decomposition (EVD) on the normalized kernel and view the values of its principal eigenvectors (modulo the possibility of
 ``higher harmonics", see text) as the intrinsic variables of the given observations.

@@

\section{Laplacian Pyramids for Data Extension} \label{sec:LapPyr}

In this work, we are not only interested in extracting the underlying variables $\mathbf{x}$ from some (partial) observations $\mathbf{Y}$,
but also interested in extending high-dimensional functions on a set of points which lie in a low-dimensional space.
%
More specifically, thinking of the ambient space coordinates, $\mathbf{Y}$, as functions on the data,
we want to estimate $\mathbf{Y}$ for new points $\mathbf{x}$.
%
Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $\mathbf{f}$ defined on a set of points.
%
The algorithm uses Laplacian kernels of decreasing widths to create multiscale representations of $\mathbf{f}$;
these representations can be easily extended to new data points.
%
This type of multiscale representation was introduced by Burt and Adelson \cite{burt1983laplacian} for image coding,
and was later shown to be a tight frame by Do and Veterli \cite{do2003framing}.
%
Recently, LP was used to extend nonlinear embedding coordinates to new high-dimensional data points \cite{rabin2012heterogeneous}.
%
%%%RONEN
% this is colloquially correct but inconsistent
%
We will first review the LP algorithm for approximating and extending a one-dimensional function,
and then describe the application of LP in extending high-dimensional functions.

Let $\mathbf{f}: \Gamma \rightarrow \mathbb{R}^n$ be a function that is known on a subset of points $S \subset \Gamma$.
%
A coarse representation of $\mathbf{f}$ is generated using a coarse smoothing operator $P_0$.
%
The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by 
\begin{equation}
p_0(i, j)= s_0^{-1}(t)w_0(i, j),
\end{equation}
where $w_0(i, j)=e^{-d^2(i, j) / \sigma_0}$ and $s_0(t)=\sum_{j \in S}w_0(i, j)$ is the normalizing term.
%
The pairwise distance $d(i, j)$ is typically the Euclidean distance, and the parameter $\sigma_0$ is set to be large compared to the values of $d^2(i, j)$.
%
The application of $P_0$ to $\mathbf{f}$ yields a coarse representation of the function, which we denote by $\mathbf{f}_0=P_0(\mathbf{f})$.
